Generalizations of Quotient Inductive-Inductive Types\thanks{This work was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST Action EUTypes CA15123.}

### 2019-10-16 draft/paper.tex
progress

@@

We also develop semantics. We show that for each signature, there is a cwf
(category with families) of algebras, extended with $\Sigma$-types, extensional
identity, and constant families. This additional structure corresponds to a
type-theoretic flavor of finite limits, and it was shown in \cite{TODO} that the
category of such cwfs is biequivalent to the category of finitely complete
categories.

@@

In Section \ref{sec:levels} we set up cumulative universes which we use in later
sections. In Section \ref{sec:tos} we introduce the theory of QIIT
signatures. In Section \ref{sec:cwfmodel} we develop initial algebra semantics,
also covering the theory of signatures itself. In Sections
\ref{sec:infinitary_term}-\ref{sec:sort_term} we present the two term model
constructions. We discuss related work and conclusions in Sections
\ref{sec:related}-\ref{sec:conclusion}.

@@

From now on, we refer to this theory as cETT (cumulative extensional type
theory). When working in cETT, we use the following notation. We
have Russell-style universes $\Set_i$ indexed by natural numbers, dependent
functions as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with
projections $\proj1$ and $\proj2$. We also use a field projection notation
for iterated pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i)
\times (f : A \ra B)$, we use $t.B$ to denote the projection of the second
component.

@@

We also need to quantify over universe levels. cETT does not support this, and
we leave it like that. Instead, for the sake of simplicity, we quantify over
levels in an unspecified metatheory \emph{outside} cETT. Hence, a universe
polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We use the $\forall$ symbol for this, e.g.\ as $\forall\,i.\,\Set_i \ra
\Set_i$. This is sufficient, because we do not need to internalize
level-polymorphic constructions in cETT.

@@

\begin{nidefinition}{Notion of model for ToS.}\label{def:tos}
For levels $i$ and $j$, there is a cETT type $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$, whose elements are
ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated $\Sigma$-type, containing
all of the following components.
\begin{enumerate}
\item
  A category with families (cwf), where all four underlying sets (of
  objects, morphisms, types and terms) are in $\Set_i$. Following notation in
  \cite{TODO}, we denote these respectively as $\Con : \Set_i$, $\Sub : \Con \ra
  \Con \ra \Set_i$, $\Ty : \Con \ra \Set_i$ and $\Tm : (\Gamma : \Con) \ra
  \Ty\,\Gamma \ra \Set_i$. We denote the empty context as $\emptycon : \Con$,
  context extension as $\blank\ext\blank : (\Gamma : \Con)\ra \Ty\,\Gamma \ra \Con$.
  Substitution on types and terms is written as $\blank[\blank]$.
\item
  A universe $\U : \Ty\,\Gamma$ with decoding $\El : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,\Gamma$.
\item
  Inductive function space, with $\Pii : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,(\Gamma\,\ext\,\El\,a) \ra \Ty\,\Gamma$, application as $\appi :
  \Tm\,\Gamma\,(\Pii\,a\,B)\ra \Tm\,(\Gamma\,\ext\,\El\,a)\,B$, and its inverse
  $\lami$.
\item
  External function space with $\Set_j$ domain: $\Pie : (A : \Set_j)\ra(A \ra
  \Ty\,\Gamma)\ra \Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra(x :
  A)\ra \Tm\,\Gamma\,(B\,x)$, and likewise $\lame$ as inverse.
\item
  Infinitary function space, with $\Piinf : (A : \Set_j)\ra(A \ra
  \Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, $\appinf :
  \Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra(x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x))$, and
  $\laminf$.
\item
  An identity type $\Id : (a : \Tm\,\Gamma\,\U)\ra
  \Tm\,\Gamma\,(\El\,a)\ra\Tm\,\Gamma\,(\El\,a)\ra \Tm\,\Gamma\,\U$, with
  $\Refl : (T : \Tm\,\Gamma\,(\El\,a))\ra \Id\,a\,t\,t$ and equality reflection.
\item
  An identity type $\IdU : \Tm\,\Gamma\,\U \ra \Tm\,\Gamma\,\U \ra \Ty\,\Gamma$, with
  evident $\ReflU$ and equality reflection.
\end{enumerate}
\end{nidefinition}
In the above listing, we omit most cwf components and equations for substitution and
$\beta\eta$-conversion, but these should be understood to be also part of $\ToS_{i,j}$.

%% \paragraph{Notational conventions.}
\emph{Notational conventions.} Cwf components by default support de Bruijn
indices, which are not easily readable. We use instead a nameful notation for
binders in context extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon \ext a :
U \ext t : \El\,a)$. We also define a type-theoretic version of $\appi$ for
convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[id, u]
\end{alignat*}

@@

\begin{example}
The signature for natural
numbers is the following, where $\emptycon$, $\blank\ext\blank$, $\U$, $\El$ and
$\Pii$ all refer to components of $M$:
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (M : \ToS_{i,0}). (\emptycon\,\ext\, N : \U \,\ext\, zero : \El\,N \ext suc : \Pii\,(n : N)\,(\El\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Recursive equations}. Our definition of $\Id$ returns in $\U$,
which, similarly to the case of infinitary functions, allows us to write equations
in $\Pii$ domains. A minimal example:
\[
RecEqSig :\equiv \emptycon \ext A : \U \ext a : \El\,A \ext f : \Pii\,(x : A)\,(\Pii\,(\Id\,A\,x\,a)\,(\El\,A))
\]
There more complicated and interesting examples, e.g.\ boundary conditions in
various cubical type theories\cite{TODO}. Note that our $\Id$ allows iterated
equations as well, but these are all made trivial in the semantics, where we
assume uniqueness of identity proofs.
\end{example}

\subsection{A Signature for the Theory of Signatures}

\todo{put this in infinitary term models beginning}

Starting from Section \ref{sec:infinitary_term}, we need to do proofs and
constructions by induction on signatures. However, the current signatures do not
support induction, analogously to how induction is not available for Church
encodings. We would like to have two things: a notion of signature induction,
and a model of ToS which supports induction. Then, we can use contexts in the
syntax of ToS (i.e.\ in the inductive model) as signatures. We provide a notion
of induction on signatures by
\begin{enumerate}
  \item Showing that there is a signature which describes ToS.
  \item Building categorical semantics, which for each signature provides a notion of induction.
\end{enumerate}
On the other hand, in this paper we will only \emph{assume} the existence of a
syntax for ToS starting from Section \ref{sec:infinitary_term}. We do not provide a
construction of the syntax from simpler type formers, although we do show that
lower-level ToS syntaxes are constructible from higher-level ones.

\section{Categories with Families of Algebras}
\label{sec:cwfmodel}

\section{Infinitary Term Models}
\label{sec:infinitary_term}

@@

\section{Term Models for Sort Equations}
\label{sec:sort_term}

>>>>

\newcommand{\Sg}{\mathsf{\Sigma}}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}

@@

We also develop semantics. We show that for each signature, there is a CwF
(category with families) of algebras, extended with $\Sigma$-types, extensional
identity, and constant families. This additional structure corresponds to a
type-theoretic flavor of finite limits, and it was shown in \cite{TODO} that the
category of such CwFs is biequivalent to the category of finitely complete
categories.

@@

In Section \ref{sec:levels} we set up cumulative universes which we use in later
sections. In Section \ref{sec:tos} we introduce the theory of QIIT
signatures. In Section \ref{sec:categorical_semantics} we develop initial algebra semantics,
also covering the theory of signatures itself. In Sections
\ref{sec:infinitary_term_model}-\ref{sec:sort_term} we present the two term model
constructions. We discuss related work and conclusions in Sections
\ref{sec:related}-\ref{sec:conclusion}.

@@

From now on, we refer to this theory as cETT (cumulative extensional type
theory). When working in cETT, we use the following notation. We
have Russell-style universes $\Set_i$ indexed by natural numbers, dependent
functions as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with
projections $\proj1$ and $\proj2$. We also use a field projection notation
for iterated pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i)
\times (f : A \ra B)$, we use $t.B$ to denote the projection of the second
component. Sometimes we omit the $t.$ prefix if it is clear from context.

@@

We also need to quantify over universe levels. cETT does not support this, and
we leave it like that. Instead, for the sake of simplicity, we quantify over
levels in an unspecified metatheory \emph{outside} cETT. Hence, a universe
polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We use the $\forall$ symbol for this, e.g.\ as $\forall\,i.\,\Set_i \ra
\Set_i$. This is sufficient in this paper, because we do not need to internalize
level-polymorphic constructions in cETT.

@@

\begin{nidefinition}{Notion of model for ToS.}\label{def:tos}
For levels $i$ and $j$, there is a cETT type $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$, whose elements are
ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated $\Sigma$-type, containing
all of the following components.
\begin{enumerate}
\item
  A category with families (CwF), where all four underlying sets (of
  objects, morphisms, types and terms) are in $\Set_i$. Following notation in
  \cite{TODO}, we denote these respectively as $\Con : \Set_i$, $\Sub : \Con \ra
  \Con \ra \Set_i$, $\Ty : \Con \ra \Set_i$ and $\Tm : (\Gamma : \Con) \ra
  \Ty\,\Gamma \ra \Set_i$. We denote the empty context as $\emptycon : \Con$,
  context extension as $\blank\ext\blank : (\Gamma : \Con)\ra \Ty\,\Gamma \ra \Con$.
  Substitution on types and terms is written as $\blank[\blank]$.
\item
  A universe $\U : \Ty\,\Gamma$ with decoding $\El : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,\Gamma$.
\item
  Inductive function space, with $\Pii : (a : \Tm\,\Gamma\,\U) \ra
  \Ty\,(\Gamma\,\ext\,\El\,a) \ra \Ty\,\Gamma$, application as $\appi :
  \Tm\,\Gamma\,(\Pii\,a\,B)\ra \Tm\,(\Gamma\,\ext\,\El\,a)\,B$, and its inverse
  $\lami$.
\item
  External function space with $\Set_j$ domain: $\Pie : (A : \Set_j)\ra(A \ra
  \Ty\,\Gamma)\ra \Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra(x :
  A)\ra \Tm\,\Gamma\,(B\,x)$, and likewise $\lame$ as inverse.
\item
  Infinitary function space, with $\Piinf : (A : \Set_j)\ra(A \ra
  \Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, $\appinf :
  \Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra(x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x))$, and
  $\laminf$.
\item
  An identity type $\Id : (a : \Tm\,\Gamma\,\U)\ra
  \Tm\,\Gamma\,(\El\,a)\ra\Tm\,\Gamma\,(\El\,a)\ra \Tm\,\Gamma\,\U$, with
  $\Refl : (T : \Tm\,\Gamma\,(\El\,a))\ra \Id\,a\,t\,t$, equality reflection
  and uniqueness of identity proofs.
\item
  An identity type $\IdU : \Tm\,\Gamma\,\U \ra \Tm\,\Gamma\,\U \ra \Ty\,\Gamma$, with
  evident $\ReflU$, equality reflection and uniqueness of identity proofs.
\end{enumerate}
\end{nidefinition}
In the above listing, we omit most CwF components and equations for substitution and
$\beta\eta$-conversion, but these should be understood to be also part of $\ToS_{i,j}$.

%% \paragraph{Notational conventions.}
\emph{Notational conventions.} CwF components by default support de Bruijn
indices, which are not easily readable. We use instead a nameful notation for
binders in context extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon \ext a :
U \ext t : \El\,a)$. We also define a type-theoretic version of $\appi$ for
convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[id, u]
\end{alignat*}

@@

\begin{example}
The signature for natural
numbers is the following, where $\emptycon$, $\blank\ext\blank$, $\U$, $\El$ and
$\Pii$ all refer to components of the $\lambda$-bound $M$:
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (M : \ToS_{i,0}). (\emptycon\,\ext\, N : \U \,\ext\, zero : \El\,N \ext suc : \Pii\,(n : N)\,(\El\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Recursive equations}. Our definition of $\Id$ returns in $\U$,
which, similarly to the case of infinitary functions, allows us to write equations
in $\Pii$ domains. A minimal example:
\[
RecEqSig :\equiv \emptycon \ext A : \U \ext a : \El\,A \ext f : \Pii\,(x : A)\,(\Pii\,(\Id\,A\,x\,a)\,(\El\,A))
\]
There are more complicated and interesting examples, e.g.\ boundary conditions in
various cubical type theories\cite{TODO}. Note that our $\Id$ allows iterated
equations as well, but these are all made trivial in the semantics, where we
assume uniqueness of identity proofs.
\end{example}

\section{Categorical Semantics}
\label{sec:categorical_semantics}

For each signature, we would like to have at least
\begin{enumerate}
  \item A category of algebras, with homomorphisms as morphisms.
  \item A notion of induction, which requires a notion of dependent algebras.
  \item A proof that for algebras, initiality is equivalent to supporting induction.
\end{enumerate}

Following \cite{TODO}, we do this by creating a model of ToS, where contexts are
categories supporting the above requirements and substitutions are strictly
structure-preserving functors. Then, each signature can applied to this model,
yielding an interpretation of the signature as a structured category of
algebras. First, we fix our notion of such structured category.

\begin{nidefinition}{Finite limit CwFs.}
For a level $i$, there is a cETT type $\flCwF_i : \Set_{i+1}$, which is an
iterated $\Sigma$-type with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with usual projections and pairing.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with equality reflection and uniqueness
    of identity proofs.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, and an isomorphism
    between $\Sub\,\Gamma\,\Delta$ and $\Tm\,\Gamma\,(\Kfam\,\Delta)$ which is
    natural in $\Gamma$. The idea is that $\Kfam\,\Delta$ is a representation
    of $\Delta$ as a type in any context. Dybjer and Clairambault called constant
    families ``democracy'' in \cite{TODO}.
\end{enumerate}
\end{nidefinition}

We use $\Ty$ in an flCwF to represent ``dependent objects'', and $\Tm$
for ``dependent morphisms''. We use these to define induction.

\begin{nidefinition} Notion of induction internally to $C : \flCwF_i$:
\begin{alignat*}{3}
  & Inductive : C.\Con \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : C.\Ty\,\Gamma)\ra C.\Tm\,\Gamma\,A
\end{alignat*}
\end{nidefinition}

In the categorical semantics for a concrete signature, this informally means
that an algebra supports induction if for any dependent algebra over it (which
is a bundle of induction motives and methods), there is a dependent morphism
into it (which is a bundle of eliminator functions and their $\beta$-rules).

\begin{theorem}{Equivalence of initiality and induction.}
An object $\Gamma : C.\Con$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof irrelevant
predicates on objects.
\end{theorem}
\begin{proof} Shown by brief internal flCwF reasoning in \cite{TODO}.\qed
\end{proof}

The reason for the ``finite limit CwF'' naming is the following: Dybjer and
Clairambault showed\cite{TODO} that the 2-category of flCwFs is biequivalent to
the 2-category of finitely complete (or ``lex'') categories. It can be also seen
that the categorical product of $\Gamma$ and $\Delta$ in an flCwF can be given
as $\Gamma \ext \Kfam\, \Delta$, and equalizers can be given using the identity
type.

So, why do we use flCwF instead of lex categories? After all, the former
structure has more components and has more immediate complexity than the
latter. The reason is that flCwFs allow us to define induction in strictly the
same way as one would write it down in type theory, since we have $\Ty$ and
$\Tm$ for a primitive notion of dependent objects and morphisms. In contrast,
dependent objects in lex categories is a derived notion, and the induction
principles we get are what we want only up to isomorphisms. This issue is
perhaps not very relevant to those with a purely categorical perspective, but we
are concerned with eventually implementing QIITs in proof assistants, so we
prefer if our semantics computes notions strictly.

\begin{nidefinition}{flCwF model for ToS.} For $i$ and $j$ levels, there is
an $M : \ToS_{1+\max(i+1,j),j}$ such that $M.\Con = \flCwF_{\max(i+1,j)}$.

\todo{flip level indices in ToS, maybe shorten cETT part, add more explanation
      on role of i,j in ToS(i,j)}

\end{nidefinition}

\section{Infinitary Term Models}
\label{sec:infinitary_term_models}

\subsection{Induction on Signatures}

\todo{sync}

Starting from Section \ref{sec:infinitary_term_models}, we need to do proofs and
constructions by induction on signatures. However, the current signatures do not
support induction, analogously to how induction is not available for Church
encodings. We would like to have two things: a notion of signature induction,
and a model of ToS which supports induction. Then, we can use contexts in the
syntax of ToS (i.e.\ in the inductive model) as signatures. We provide a notion
of induction on signatures by
\begin{enumerate}
  \item Showing that there is a signature which describes ToS.
  \item Building categorical semantics, which for each signature provides a notion of induction.
\end{enumerate}
On the other hand, in this paper we will only \emph{assume} the existence of a
syntax for ToS starting from Section \ref{sec:infinitary_term_models}. We do not provide a
construction of the syntax from simpler type formers, although we do show that
lower-level ToS syntaxes are constructible from higher-level ones.

@@

\section{Term Models with Sort Equations}
\label{sec:sort_term}
